---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 10,
  fig.height = 7
)
```

# Global explanations for tree-based models

<!-- badges: start -->
[![R-CMD-check](https://github.com/PlantedML/glex/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PlantedML/glex/actions/workflows/R-CMD-check.yaml)
[![glex status badge](https://plantedml.r-universe.dev/glex/badges/version)](https://plantedml.r-universe.dev/glex)
[![CRAN status](https://www.r-pkg.org/badges/version/glex)](https://CRAN.R-project.org/package=glex)
<!-- badges: end -->

# Overview

`glex` implements **global, functional decompositions** of tree-based regression and classification models that decomposes them into **main effects** and **interaction effects** of arbitrary order. In particular, it can

- compute **exact interventional SHAP values** (and more generally, *q-interaction interventional SHAP*, where _q_ is the maximal interaction order present in the model),
- extract **partial-dependence-functions** (e.g., one dimensional partial dependence plots),
- produce **variable‐importance scores** for each main term and interaction term (of any order), and
- supports **de-biasing** by removing components that include protected features.

Under the hood, `glex` relies on fast algorithms to compute all required partial dependence functions exactly. 

For a detailed description of the methodology, see:

See the accompanying paper for more details and exact definitions:
"Unifying local and global model explanations by functional
decomposition of low dimensional structures"
([arxiv](https://arxiv.org/abs/2208.06151), [PMLR](https://proceedings.mlr.press/v206/hiabu23a.html)).

> **Hiabu, Meyer & Wright (2023).**  
> _Unifying local and global model explanations by functional decomposition of low dimensional structures._  
> [arXiv](https://arxiv.org/abs/2208.06151) • [AISTATS 2023 Proceedings](https://proceedings.mlr.press/v206/hiabu23a.html)

> **Liu, Steensgaard, Wright, Pfister, Hiabu (2023).**  
> _Fast Estimation of Partial Dependence Functions using Trees._  
> [arXiv](https://arxiv.org/abs/2410.13448) 


## Installation

You can install the development version of glex from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("PlantedML/glex")
```

or from [r-universe][r-universe] with

```r
install.packages("glex", repos = "https://plantedml.r-universe.dev")
```

# Supported Models

`glex` currently provides methods for the model classes below.

| Model package | Model class | Regression | Binary classification | Multiclass classification | Link function(s) | Notes |
|---|---|---|---|---|---|---|
| `xgboost` | `xgb.Booster` | Yes | Yes* | Not yet fully supported | Built-in objectives define the link (e.g., identity, logistic/logit, log-link). | * `x` must be a numeric matrix. `glex()` decomposes predictions on the raw margin scale; apply the inverse link to recover response-scale predictions. |
| `randomPlantedForest` | `rpf` | Yes | Yes | Yes | Not applicable | Native support for multiclass terms in plotting and variable importance workflows. |
| `ranger` | `ranger` | Yes | Yes* (probability forests) | Not yet supported | Not applicable | * Requires `node.stats = TRUE`. For classification, fit with `probability = TRUE`; ranger predicts class probabilities directly from class frequencies in terminal nodes (no inverse link needed). Multiclass is currently unsupported. |

More tree-based frameworks may be added in future releases. If you have a suggestion, please open an issue on our GitHub repository.

### XGBoost with Link Functions

For generalized objectives, XGBoost outputs an additive raw-score (margin) function:

$$
\eta(x) = b + \sum_{t=1}^{T} f_t(x), \qquad \mu(x)=g^{-1}(\eta(x)).
$$

Here, $b$ is the global bias term (`base_score` on the margin scale), and each $f_t$ is the prediction function of tree $t$ (its leaf weight for input $x$). In other words, the XGBoost model output is $\eta(x)$ itself; response-scale prediction is obtained by applying the objective-specific inverse link $g^{-1}$ to that output.  
`glex()` decomposes $\eta(x)$, not $\mu(x)$. The decomposition is

$$
\eta(x) = m_{\emptyset} + \sum_{S \neq \emptyset} m_S(x_S) = m_{\emptyset} + \sum_{k=1}^{p} \phi_k(x),
$$

where $m_{\emptyset}$ is the intercept term, $m_S$ are the functional ANOVA components indexed by feature subsets $S$, and $\phi_k$ are SHAP values aggregated per feature $k$. This margin equals `predict(model, x, outputmargin = TRUE)`.  
Predictions on response scale are obtained by applying the inverse link:

$$
\mu(x)=g^{-1}(\eta(x))=g^{-1}\left(m_{\emptyset} + \sum_{S \neq \emptyset} m_S(x_S)\right).
$$

This yields the objective-specific identities:

$$
\text{identity link:}\quad \mu(x)=\eta(x)
$$

$$
\text{logistic link (binary:logistic):}\quad \mu(x)=\sigma(\eta(x))=\frac{1}{1+e^{-\eta(x)}}
$$

$$
\text{log link (count:poisson, reg:gamma, reg:tweedie):}\quad \mu(x)=\exp(\eta(x))
$$

For custom objectives, XGBoost does not provide a built-in response transform. In that case, run `glex()` as usual to decompose the raw margin, then apply your own inverse link **afterwards** to the reconstructed margin, e.g. to `gl$intercept + rowSums(gl$m)` (or equivalently `gl$intercept + rowSums(gl$shap)`), to obtain response-scale predictions.

```{r xgb-link-functions}
x <- as.matrix(mtcars[, -1])
y_bin <- as.numeric(mtcars$mpg > median(mtcars$mpg))

xg_bin <- xgboost::xgb.train(
  params = list(objective = "binary:logistic", max_depth = 3, eta = .1),
  data = xgboost::xgb.DMatrix(data = x[1:26, ], label = y_bin[1:26]),
  nrounds = 30,
  verbose = 0
)

glex_xgb_bin <- glex::glex(xg_bin, x[27:32, ])

# Additive decomposition returned by glex (margin / log-odds scale)
margin_glex <- glex_xgb_bin$intercept + rowSums(glex_xgb_bin$shap)

# XGBoost predictions on test data
pred_prob <- predict(xg_bin, x[27:32, ])
pred_margin <- predict(xg_bin, x[27:32, ], outputmargin = TRUE)

# Apply the inverse link for binary:logistic
prob_from_glex <- plogis(margin_glex)

cbind(pred_prob, prob_from_glex, pred_margin, margin_glex)

max(abs(pred_margin - margin_glex))
max(abs(pred_prob - prob_from_glex))
```

Binary logistic example (`objective = "binary:logistic"`):

In this case $\eta(x)$ is log-odds and the inverse link is the logistic map:

$$
F(x) = b + \sum_{t=1}^{T} f_t(x), \qquad p(x) = \sigma(F(x)) = \frac{1}{1 + e^{-F(x)}}.
$$

XGBoost optimizes logistic loss directly in this margin:

$$
\ell(y, F) = -\left[y \log \sigma(F) + (1-y)\log(1-\sigma(F))\right].
$$

`glex()` decomposes the margin additively into interaction components indexed by feature subsets $S$:

$$
F(x) = m_{\emptyset} + \sum_{S \neq \emptyset} m_S(x_S),
$$

where `intercept` is $m_{\emptyset}$ and `m` stores $m_S$. Hence:

$$
\texttt{intercept + rowSums(m)} \equiv \texttt{predict(..., outputmargin = TRUE)}.
$$

For SHAP values, `glex` distributes each interaction term equally across features in that term:

$$
\phi_j(x) = \sum_{S \ni j} \frac{m_S(x_S)}{|S|},
\qquad
F(x) = m_{\emptyset} + \sum_{j=1}^{p}\phi_j(x).
$$

Hence:

$$
\texttt{plogis(intercept + rowSums(shap))} \equiv \texttt{predict(...)}.
$$

Interpretation on this scale is direct: positive components increase log-odds (and therefore probability), negative components decrease log-odds, and an increase of $+1$ in margin multiplies the odds by $e$.

## What's Included

The examples below use [`xgboost`][xgb] and [`randomPlantedForest`][rpf] models:

```r
# Install xgboost from CRAN
install.packages("xgboost")
# ... and randomPlantedForest from r-universe
install.packages("randomPlantedForest", repos = "https://plantedml.r-universe.dev")
```

```{r setup}
library(glex)

# Model fitting
library(randomPlantedForest)
library(xgboost)

# Visualization
library(ggplot2)
library(patchwork)

theme_set(theme_glex())
set.seed(21)
```

Note that `xgboost`, unlike `randomPlantedForest`, requires `matrix` input and does not support categorical predictors.

```{r fit-models}
rp <- rpf(mpg ~ ., data = mtcars[1:26, ], max_interaction = 3)

x <- as.matrix(mtcars[, -1])
y <- mtcars$mpg
xg <- xgboost(data = x[1:26, ], label = y[1:26],
              params = list(max_depth = 3, eta = .1),
              nrounds = 30, verbose = 0)

```

Using the model objects and a dataset to explain (such as a test set in this case), we can create `glex` objects
for each of the model fits.
These objects of class `glex` are a list containing the prediction components of main and interaction terms (`$m`), the dataset to be explained with
the observed feature values (`$x`) used to visualize feature effects, and the average predicted value for the model (`$intercept`).
The `xgboost` method additionally returns the SHAP values (`$shap`) for each feature in the model.

```{r glex}
glex_rpf <- glex(rp, mtcars[27:32, ])
glex_xgb <- glex(xg, x[27:32, ])
```

Both `m` and `shap` satisfy the property that their sums (per observation) together with the `intercept` are equal to the model prediction for each observation:

```{r comp-sum}
# Calculating sum of components and sum of SHAP values
sum_m_rpf <- rowSums(glex_rpf$m) + glex_rpf$intercept
sum_m_xgb <- rowSums(glex_xgb$m) + glex_xgb$intercept
sum_shap_xgb <- rowSums(glex_xgb$shap) + glex_xgb$intercept

# Model predictions
pred_rpf <- predict(rp, mtcars[27:32, ])[[1]]
pred_xgb <- predict(xg, x[27:32, ])

# For XGBoost
cbind(pred_xgb, sum_m_xgb, sum_shap_xgb)

# For RPF
cbind(pred_rpf, sum_m_rpf)
```

### Variable Importances

Variable importance scores are calculated for each main and interaction term by calculating the average of the absolute prediction components (`m`) over the dataset supplied to `glex()`.

```{r glex_vi}
vi_rpf <- glex_vi(glex_rpf)
vi_xgb <- glex_vi(glex_xgb)

vi_rpf[1:5, c("degree", "term", "m")]
vi_xgb[1:5, c("degree", "term", "m")]
```

The output additionally contains the degree of interaction, which can be used for filtering and aggregating. Here we filter for terms with contributions above a `threshold` of `0.05` to get a more compact plot, with terms below the threshold aggregated into one labelled "Remaining terms":

```{r glex_vi-plot, fig.width=12}
p_vi1 <- autoplot(vi_rpf, threshold = .05) + 
  labs(title = NULL, subtitle = "RPF")

p_vi2 <- autoplot(vi_xgb, threshold = .05) + 
  labs(title = NULL, subtitle = "XGBoost")

p_vi1 + p_vi2 +
  plot_annotation(title = "Variable importance scores by term")
```

We can also sum values within each degree of interaction for a more aggregated view, which can be useful as it allows us to judge interactions above a certain degree to not be particularly relevant for a given model.

```{r glex_vi-plot-by-degree, fig.height=5}
p_vi1 <- autoplot(vi_rpf, by_degree = TRUE) + 
  labs(title = NULL, subtitle = "RPF")

p_vi2 <- autoplot(vi_xgb, by_degree = TRUE) + 
  labs(title = NULL, subtitle = "XGBoost")

p_vi1 + p_vi2 +
  plot_annotation(title = "Variable importance scores by degree") 
```

### Feature Effects

We can also plot prediction components against observed feature values, which admittedly produces more interesting output with larger, more interesting datasets.

```{r plot_components}
p1 <- autoplot(glex_rpf, "hp") + labs(subtitle = "RPF")
p2 <- autoplot(glex_xgb, "hp") + labs(subtitle = "XGBoost")

p1 + p2 + 
  plot_annotation(title = "Main effect for 'hp'")

p1 <- autoplot(glex_rpf, c("hp", "wt")) + labs(subtitle = "RPF")
p2 <- autoplot(glex_xgb, c("hp", "wt")) + labs(subtitle = "XGBoost")

p1 + p2 + 
  plot_annotation(title = "Two-way effects for 'hp' and 'wt'")
```

Currently there is support for plots of interactions up to the third degree, including continuous and categorical features. Unfortunately, three-way interactions of continuous features are not supported yet.

Note that these main effect plots correspond to PDP plots, where the latter are merely the main effect plus the intercept term:

```{r main_pdp}
plot_pdp(glex_rpf, "hp")
```

### Decomposition of Individual Predictions

Finally, we can explore the prediction for a single observation by displaying its individual prediction components. The SHAP value is the sum of all of these components and serves as a reference value. 
For compactness, we only plot one feature and collapse all interaction terms above the second degree into one as their combined effect is very small.

```{r glex_explain}
p1 <- glex_explain(glex_rpf, id = 2, predictors = "hp", max_interaction = 2) + 
  labs(tag = "RPF")
p2 <- glex_explain(glex_xgb, id = 2, predictors = "hp", max_interaction = 2) + 
  labs(tag = "XGBoost")

p1 + p2 & theme(plot.tag.position = "bottom")
```

[xgb]: https://cran.r-project.org/package=xgboost
[rpf]: http://plantedml.com/randomPlantedForest/
[r-universe]: https://plantedml.r-universe.dev/packages
