---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 10,
  fig.height = 7
)
```

# Global explanations for tree-based models

<!-- badges: start -->
[![R-CMD-check](https://github.com/PlantedML/glex/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PlantedML/glex/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/PlantedML/glex/branch/master/graph/badge.svg)](https://app.codecov.io/gh/PlantedML/glex)
[![glex r-universe status badge](https://plantedml.r-universe.dev/badges/glex)](https://plantedml.r-universe.dev/glex)
[![CRAN status](https://www.r-pkg.org/badges/version/glex)](https://CRAN.R-project.org/package=glex)
<!-- badges: end -->

Global explanations for tree-based models by decomposing regression or  
classification functions into the sum of main components and interaction
components of arbitrary order. Calculates exact interventional SHAP
values and q-interaction interventional SHAP (q being the highest order
of interaction present in the model) for tree-based models such as
XGBoost.

See the accompanying paper for more details and exact definitions:
"Unifying local and global model explanations by functional
decomposition of low dimensional structures"
([arxiv](https://arxiv.org/abs/2208.06151), [PMLR](https://proceedings.mlr.press/v206/hiabu23a.html)).

## Installation

You can install the development version of glex from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("PlantedML/glex")
```

or from [r-universe][r-universe] with

```r
install.packages("glex", repos = "https://plantedml.r-universe.dev")
```

## What's Included

Currently `glex` works with [`xgboost`][xgb] and [`randomPlantedForest`][rpf] models, so we'll start by fitting a model with one of each:

```r
# Install xgboost from CRAN
install.packages("xgboost")
# ... and randomPlantedForest from r-universe
install.packages("randomPlantedForest", repos = "https://plantedml.r-universe.dev")
```

```{r setup}
library(glex)

# Model fitting
library(randomPlantedForest)
library(xgboost)

# Visualization
library(ggplot2)
library(patchwork)

theme_set(theme_glex())
set.seed(21)
```

Note that `xgboost`, unlike `randomPlantedForest`, requires `matrix` input and does not support categorical predictors.

```{r fit-models}
rp <- rpf(mpg ~ ., data = mtcars[1:26, ], max_interaction = 3)

x <- as.matrix(mtcars[, -1])
y <- mtcars$mpg
xg <- xgboost(data = x[1:26, ], label = y[1:26],
              params = list(max_depth = 3, eta = .1),
              nrounds = 30, verbose = 0)

```

Using the model objects and a dataset to explain (such as a test set in this case), we can create `glex` objects
for each of the model fits.
These objects of class `glex` are a list containing the prediction components of main and interaction terms (`$m`), the dataset to be explained with
the observed feature values (`$x`) used to visualize feature effects, and the average predicted value for the model (`$intercept`).
The `xgboost` method additionally returns the SHAP values (`$shap`) for each feature in the model.

```{r glex}
glex_rpf <- glex(rp, mtcars[27:32, ])
glex_xgb <- glex(xg, x[27:32, ])
```

Both `m` and `shap` satisfy the property that their sums (per observation) together with the `intercept` are equal to the model prediction for each observation:

```{r comp-sum}
# Calculating sum of components and sum of SHAP values
sum_m_rpf <- rowSums(glex_rpf$m) + glex_rpf$intercept
sum_m_xgb <- rowSums(glex_xgb$m) + glex_xgb$intercept
sum_shap_xgb <- rowSums(glex_xgb$shap) + glex_xgb$intercept

# Model predictions
pred_rpf <- predict(rp, mtcars[27:32, ])[[1]]
pred_xgb <- predict(xg, x[27:32, ])

# For XGBoost
cbind(pred_xgb, sum_m_xgb, sum_shap_xgb)

# For RPF
cbind(pred_rpf, sum_m_rpf)
```


### Variable Importances

Variable importance scores are calculated for each main and interaction term by calculating the average of the absolute prediction components (`m`) over the dataset supplied to `glex()`.

```{r glex_vi}
vi_rpf <- glex_vi(glex_rpf)
vi_xgb <- glex_vi(glex_xgb)

vi_rpf[1:5, c("degree", "term", "m")]
vi_xgb[1:5, c("degree", "term", "m")]
```

The output additionally contains the degree of interaction, which can be used for filtering and aggregating. Here we filter for terms with contributions above a `threshold` of `0.05` to get a more compact plot, with terms below the threshold aggregated into one labelled "Remaining terms":

```{r glex_vi-plot, fig.width=12}
p_vi1 <- autoplot(vi_rpf, threshold = .05) + 
  labs(title = NULL, subtitle = "RPF")

p_vi2 <- autoplot(vi_xgb, threshold = .05) + 
  labs(title = NULL, subtitle = "XGBoost")

p_vi1 + p_vi2 +
  plot_annotation(title = "Variable importance scores by term")
```

We can also sum values within each degree of interaction for a more aggregated view, which can be useful as it allows us to judge interactions above a certain degree to not be particularly relevant for a given model.

```{r glex_vi-plot-by-degree, fig.height=5}
p_vi1 <- autoplot(vi_rpf, by_degree = TRUE) + 
  labs(title = NULL, subtitle = "RPF")

p_vi2 <- autoplot(vi_xgb, by_degree = TRUE) + 
  labs(title = NULL, subtitle = "XGBoost")

p_vi1 + p_vi2 +
  plot_annotation(title = "Variable importance scores by degree") 
```

### Feature Effects

We can also plot prediction components against observed feature values, which admittedly produces more interesting output with larger, more interesting datasets.

```{r plot_components}
p1 <- autoplot(glex_rpf, "hp") + labs(subtitle = "RPF")
p2 <- autoplot(glex_xgb, "hp") + labs(subtitle = "XGBoost")

p1 + p2 + 
  plot_annotation(title = "Main effect for 'hp'")

p1 <- autoplot(glex_rpf, c("hp", "wt")) + labs(subtitle = "RPF")
p2 <- autoplot(glex_xgb, c("hp", "wt")) + labs(subtitle = "XGBoost")

p1 + p2 + 
  plot_annotation(title = "Two-way effects for 'hp' and 'wt'")
```

Currently there is support for plots of interactions up to the third degree, including continuous and categorical features. Unfortunately, three-way interactions of continuous features are not supported yet.

Note that these main effect plots correspond to PDP plots, where the latter are merely the main effect plus the intercept term:

```{r main_pdp}
plot_pdp(glex_rpf, "hp")
```

### Decomposition of Individual Predictions

Finally, we can explore the prediction for a single observation by displaying its individual prediction components. The SHAP value is the sum of all of these components and serves as a reference value. 
For compactness, we only plot one feature and collapse all interaction terms above the second degree into one as their combined effect is very small.

```{r glex_explain}
p1 <- glex_explain(glex_rpf, id = 2, predictors = "hp", max_interaction = 2) + 
  labs(tag = "RPF")
p2 <- glex_explain(glex_xgb, id = 2, predictors = "hp", max_interaction = 2) + 
  labs(tag = "XGBoost")

p1 + p2 & theme(plot.tag.position = "bottom")
```

[xgb]: https://cran.r-project.org/package=xgboost
[rpf]: http://plantedml.com/randomPlantedForest/
[r-universe]: https://plantedml.r-universe.dev/packages
