---
title: "Categorical features: native factors, one-hot groups and semantic groups"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Categorical features: native factors, one-hot groups and semantic groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_deps <- requireNamespace("xgboost", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  eval = has_deps
)
```

```{r setup, message=FALSE}
library(glex)
library(xgboost)
set.seed(1)
```

A categorical predictor can reach `xgboost` in two ways, and `glex` handles
both:

1. **Native factors** (`xgboost >= 3.0`): fit on a `data.frame` with factor
   columns. The trees split on sets of levels, and the decomposition treats
   the factor as one feature from the start.
2. **Dummy encoding** via `model.matrix()`: the classic route, and the only one
   for models fit before native support. The decomposition then sees one
   feature per level, and `group_components()` re-assembles the factor
   afterwards.

`group_components()` also serves a third purpose: treating any set of features
as one, e.g. a block of related measurements.

The built-in `bike` data has a four-level `season` factor alongside numeric
predictors:

```{r data}
data(bike, package = "glex")
x_df <- as.data.frame(bike[, c("season", "hr", "temp", "hum")])
str(x_df)
```

## Native factors

With a `data.frame` input, `xgboost()` treats factor columns as categorical.
Pass the same `data.frame` to `glex()`; the factor levels must match the
training data in order, as for `predict()`.

```{r native}
xg_native <- xgboost(
  x_df,
  bike$bikers,
  nrounds = 50,
  max_depth = 3,
  learning_rate = 0.1,
  nthreads = 2
)
res_native <- glex(xg_native, x_df)
names(res_native$m)
```

`season` is one feature: there is a single main effect and single interactions
with the numeric predictors. The factor column is kept in `$x`, so plots treat
it as categorical:

```{r native-plots, fig.height=4}
plot_main_effect(res_native, "season")
plot_twoway_effects(res_native, c("season", "temp"))
```

## Dummy encoding fragments the decomposition

Encoding `season` with `model.matrix()` gives one column per level. The
decomposition then treats every dummy column as a feature of its own: the
effect of the factor is smeared across its dummies *and* their mutual
"interactions", which are artifacts of the encoding rather than anything about
the data.

```{r dummy}
x <- cbind(
  model.matrix(~ season - 1, bike),
  as.matrix(bike[, c("hr", "temp", "hum")])
)
colnames(x)

xg <- xgb.train(
  params = xgb.params(max_depth = 3, learning_rate = 0.1, nthread = 2),
  data = xgb.DMatrix(x, label = bike$bikers, nthread = 2),
  nrounds = 50
)
res <- glex(xg, x)
ncol(res$m)
```

Among these terms are things like `seasonFall:seasonWinter`, an "interaction"
between two levels of the same factor:

```{r raw-terms}
head(sort(colMeans(abs(as.matrix(res$m))), decreasing = TRUE), 10)
```

### Re-assembling the factor with `group_components()`

`group_components()` aggregates all terms feature-wise: terms involving only
`season` dummies become the `season` main effect, terms mixing a dummy with
`hr` become the `season:hr` interaction, and so on. Because the decomposition
is additive, this is exact.

`dummy_groups()` builds the required `groups` list from the *un-encoded* data,
matching the `model.matrix()` naming convention by default. `bike` carries
further factors (`mnth`, `workingday`, `weathersit`) that we did not encode
into `x`; `dummy_groups()` reports them and moves on:

```{r group}
dummy_groups(res, bike)
grouped <- group_components(res, dummy_groups(res, bike))
names(grouped$m)
```

The components still sum to the model prediction, and the SHAP value of
`season` is the sum of its dummies' SHAP values, so the efficiency property is
preserved:

```{r exact}
all.equal(rowSums(grouped$m), rowSums(res$m))
all.equal(
  grouped$shap$season,
  rowSums(as.matrix(res$shap[, startsWith(names(res$shap), "season"), with = FALSE]))
)
```

Since the grouped columns of `x` form a one-hot encoding, `group_components()`
reconstructs the factor, and the plot functions work as for the native model:

```{r plots, fig.height=4}
plot_main_effect(grouped, "season")
plot_twoway_effects(grouped, c("season", "temp"))
```

Both routes end with `season` as one feature, but they explain different
models: native categorical splits partition the levels freely, while a dummy
split isolates one level at a time. Grouping is an exact regrouping of the
dummy model's decomposition, not an approximation of the native one.

### Other encoding schemes

If the dummy columns are not named in `model.matrix()` style, pass a `naming`
function of `(feature, levels)` that reproduces the column names the model was
trained with:

```{r naming, eval=FALSE}
dummy_groups(
  res,
  bike,
  naming = function(feature, levels) paste(feature, levels, sep = "_")
)
```

`dummy_groups()` only returns columns that actually exist in `object$x`, so a
dropped reference level (treatment coding via `~ season`) is skipped
automatically; the reconstructed factor then shows a `"(base)"` level for
observations of the reference category.

## Semantic groups

Grouping is not limited to dummy encodings. Any set of features can be treated
as one, for example a "weather" block, which in higher-dimensional settings
might be a group of anthropometrics, demographics or biomarkers:

```{r semantic}
weather <- group_components(res, groups = list(weather = c("temp", "hum")))
head(sort(colMeans(abs(as.matrix(weather$m))), decreasing = TRUE), 10)
```

For such groups no single x value exists, so `weather$x$weather` is `NA` and
the effect plots are not available for that group, but term importances and
SHAP values aggregate as usual:

```{r semantic-vi, fig.height=4}
library(ggplot2)
autoplot(glex_vi(weather), threshold = 0.5)
```
