---
title: "Grouping features: dummy-encoded factors and semantic groups"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grouping features: dummy-encoded factors and semantic groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_deps <- requireNamespace("xgboost", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  eval = has_deps
)
```

```{r setup, message=FALSE}
library(glex)
library(xgboost)
set.seed(1)
```

## The problem: dummy encoding fragments the decomposition

`xgboost` only accepts numeric features, so categorical predictors are usually
dummy-encoded with `model.matrix()` before fitting. The decomposition then
treats every dummy column as a feature of its own: the effect of one factor is
smeared across its dummies *and* their mutual "interactions", which are
artifacts of the encoding rather than anything about the data.

The built-in `bike` data has a four-level `season` factor:

```{r data}
data(bike, package = "glex")

x <- cbind(
  model.matrix(~ season - 1, bike),
  as.matrix(bike[, c("hr", "temp", "hum")])
)
colnames(x)
```

```{r fit}
xg <- xgb.train(
  params = xgb.params(max_depth = 3, learning_rate = 0.1, nthread = 2),
  data = xgb.DMatrix(x, label = bike$bikers, nthread = 2),
  nrounds = 50
)
res <- glex(xg, x)
ncol(res$m)
```

Among these terms are things like `seasonFall:seasonWinter` — an "interaction"
between two levels of the same factor, which no one wants to interpret:

```{r raw-terms}
head(sort(colMeans(abs(as.matrix(res$m))), decreasing = TRUE), 10)
```

## Re-assembling the factor with `group_components()`

`group_components()` aggregates all terms feature-wise: terms involving only
`season` dummies become the `season` main effect, terms mixing a dummy with
`hr` become the `season:hr` interaction, and so on. Because the decomposition
is additive, this is exact — no approximation involved.

`dummy_groups()` builds the required `groups` list from the *un-encoded* data,
matching the `model.matrix()` naming convention by default. `bike` carries
further factors (`mnth`, `workingday`, `weathersit`) that we did not encode
into `x` — `dummy_groups()` reports them and moves on:

```{r group}
dummy_groups(res, bike)

grouped <- group_components(res, dummy_groups(res, bike))
ncol(grouped$m)
head(sort(colMeans(abs(as.matrix(grouped$m))), decreasing = TRUE), 10)
```

The components still sum to the model prediction, and the SHAP value of
`season` is the sum of its dummies' SHAP values, so the efficiency property is
preserved:

```{r exact}
all.equal(rowSums(grouped$m), rowSums(res$m))
all.equal(
  grouped$shap$season,
  rowSums(as.matrix(res$shap[, startsWith(names(res$shap), "season"), with = FALSE]))
)
```

Since the grouped columns of `x` form a one-hot encoding, `group_components()`
reconstructs the factor, so the plot functions work on the grouped object:

```{r plots, fig.height=4}
plot_main_effect(grouped, "season")
plot_twoway_effects(grouped, c("season", "temp"))
```

## Other encoding schemes

If the dummy columns are not named in `model.matrix()` style, pass a `naming`
function of `(feature, levels)` that reproduces the column names the model was
trained with:

```{r naming, eval=FALSE}
dummy_groups(
  res,
  bike,
  naming = function(feature, levels) paste(feature, levels, sep = "_")
)
```

`dummy_groups()` only returns columns that actually exist in `object$x`, so a
dropped reference level (treatment coding via `~ season`) is skipped
automatically — the reconstructed factor then shows a `"(base)"` level for
observations of the reference category.

## Semantic groups

Grouping is not limited to dummy encodings. Any set of features can be treated
as one, for example a "weather" block:

```{r semantic}
weather <- group_components(res, groups = list(weather = c("temp", "hum")))
head(sort(colMeans(abs(as.matrix(weather$m))), decreasing = TRUE), 10)
```

For such groups no single x-value exists, so `weather$x$weather` is `NA` and
the effect plots are not available for the group's main effect — but term
importances and SHAP values aggregate as usual:

```{r semantic-vi, fig.height=4}
library(ggplot2)
autoplot(glex_vi(weather), threshold = 0.5)
```
